% opencmdb — User Manual. Build: see ../Makefile (LuaLaTeX; TEXINPUTS -> ../common).
\documentclass[11pt,oneside,listof=totoc,bibliography=totoc]{scrreprt}
\usepackage{opencmdb-manual}
\setmanualname{User Manual}

\begin{document}
\makemanualtitle{User Manual}{Understanding and operating opencmdb day to day}

\tableofcontents

\chapter{Introduction}
\productname{} is a self-hosted, single-binary engine that helps you keep an accurate,
living picture of your network. Its core idea is a comparison:

\begin{itemize}
  \item the \textbf{observed} state — what opencmdb \emph{discovers} automatically from
        your network (devices, IP addresses, switch ports, DHCP leases, \ldots);
  \item the \textbf{declared} state — what \emph{you} have documented as intended.
\end{itemize}

The difference between the two — the \textbf{gap} — is what opencmdb surfaces for you to
review. Observed data is never allowed to overwrite what you declared: the two are kept as
distinct, linked records.

\begin{note}
This manual is written for the person who \emph{uses} opencmdb — reviewing discoveries,
documenting devices, and managing IP space. Installation, configuration, security, and
backups live in the companion \textbf{Administrator Manual}.
\end{note}

\section{Who it is for}
Advanced home labs and small organisations without a dedicated IT team, typically running
on a Synology NAS. The reference scale is on the order of a few hundred hosts across a few
dozen subnets.

\section{What it is not}
opencmdb does not reconfigure your network and does not require elevated network
privileges to observe it. It documents and reconciles; it does not control.

\chapter{Core concepts}
\section{Declared vs.\ observed}
Every entity in opencmdb has (up to) two faces: a \emph{declared} record you author, and an
\emph{observed} record discovery produces. They are linked, never merged. Reconciliation
\emph{links} an observation to the declaration it matches; it never rewrites your intent.

\section{Device identity}
A device is identified by a composite identity — not by its MAC address alone — so it stays
recognisable across a changing IP or MAC. When the evidence is ambiguous, opencmdb
\textbf{abstains} rather than guessing or merging: an unresolved match is a first-class,
recorded outcome, presented with its supporting evidence.

\begin{planned}
The identity engine and reconciliation loop are the first implementation milestone. Until
they land, the behaviour described here reflects the designed model.
\end{planned}

\section{Sources and liveness}
Each discovery source reports on two independent axes: \emph{liveness}
(\texttt{live} or \texttt{blind}, with a named cause) and \emph{capability} (what it can
currently see). When a source goes blind, opencmdb freezes its last-known state and
suppresses observation-derived alerts for that scope — it never fabricates a divergence from
a source that cannot see.

\section{Triage gestures}
When you review a discovery, you choose one canonical gesture:
\begin{itemize}
  \item \textbf{Document} — close the gap by adopting the observed value into your
        declaration, field by field. You document a \emph{value}.
  \item \textbf{Accept-gap} — record that you have \emph{seen} the difference but are not
        deciding yet. The gap stays open and keeps counting; a note is required. You accept a
        \emph{gap}.
  \item \textbf{Exclude} — this item is out of scope and should stop appearing.
  \item \textbf{Attach}, \textbf{create}, \textbf{snooze} — link to an existing record,
        create a new declaration, or defer.
\end{itemize}

\begin{note}
opencmdb remembers your triage decisions (including exclusions), so a resolved item does not
reappear on the next scan.
\end{note}

\chapter{Getting started}
\begin{planned}
A guided first-run walkthrough will appear here once the application is runnable. For now,
deployment is covered in the Administrator Manual (Docker on Synology is the priority
target).
\end{planned}

\section{First light}
On first launch you will connect at least one discovery source and let opencmdb build its
initial observed picture. Nothing is alarming on day one: undocumented discoveries simply
wait for you in the triage inbox.

\chapter{Discovering your network}
\section{The UniFi source}
If you run a UniFi controller, opencmdb can discover devices, IP addresses, switch ports,
SSIDs, VLANs, and DHCP leases from it using only an API key — no elevated network access.

\section{The ARP/ping scanner}
For everything else, declare the subnets to scan and opencmdb will find active devices on
them, enriching with hostnames where available.

\begin{planned}
Both sources are designed and specified; the ARP/ping scanner is the first source to be
implemented, followed by the full UniFi connector.
\end{planned}

\chapter{The triage inbox}
The inbox is where undocumented or changed discoveries wait for your decision. It shows, in
priority order, what changed since your last visit — new devices and genuine conflicts
first, routine churn last.

\section{Documenting a discovery}
\begin{itemize}
  \item \textbf{Document-all} adopts a brand-new discovery in a single action.
  \item \textbf{Document-field} lets you adopt specific fields of a device that has drifted
        since you last documented it. The observed record itself is never modified.
\end{itemize}

\section{Working in bulk}
You can triage many discoveries at once when they share a decision, keeping steady-state
review fast.

\begin{planned}
The inbox, the document/accept-gap/exclude gestures, and bulk triage are specified in the UX
design and will be implemented over several milestones.
\end{planned}

\chapter{IP address management}
opencmdb manages subnets, VLANs, and DHCP ranges, and shows per-subnet occupancy — used,
free, declared, and observed. It can find a free address within a subnet and detect
conflicts (the same IP on two MACs, or a statically declared IP inside a DHCP range),
identifying the devices involved. IPv6 addresses can be documented (observation-only at
this stage).

\chapter{Applications and hosted software}
Beyond the network layer, you can record software instances (name, version, listening ports)
hosted on a device, and group them into applications with an owner and a criticality. This
is the ``light CMDB'' layer: enough to answer \emph{what runs where, and who owns it}.

\begin{planned}
The applications layer is planned for a later milestone, after the core reconciliation loop.
\end{planned}

\chapter{Alerts and notifications}
opencmdb can notify you of the things that matter — new conflicts, a source going blind, a
capability downgrade — without drowning you in routine change. Alerts derived from
observation are suppressed for any source that is currently blind.

\chapter{A bilingual interface}
The user interface is available in English and French; documentation is English-only. A dark
theme is the default, and key views target WCAG~2.1~AA accessibility.

\appendix
\chapter{Glossary}
\begin{description}[style=nextline,leftmargin=0pt]
  \item[Declared record] What you have documented as intended. Never overwritten by
        discovery.
  \item[Observed record] What discovery found. Linked to a declaration, never merged into it.
  \item[Gap] A difference between a declaration and its linked observation — the unit of
        work opencmdb surfaces.
  \item[Document] Close a gap by adopting an observed value into your declaration.
  \item[Accept-gap] Acknowledge a gap without resolving it; it stays open and keeps counting.
  \item[Blind source] A source that currently cannot see; its last-known state is frozen and
        its alerts suppressed.
  \item[Abstention] A recorded ``unresolved'' outcome when identity is ambiguous — never a
        guess or a merge.
\end{description}

\end{document}
